Evolutionary adaptation of the chromodomain of the HP1-protein Rhino allows the integration of chromatin and DNA sequence signals

Version 1

Abstract

Abstract

Members of the diverse heterochromatin protein 1 (HP1) family of proteins play crucial roles in heterochromatin formation and maintenance. Despite the similar affinities of their chromodomains for di- and tri-methylated histone H3 lysine 9 (H3K9me2/3), different HP1 proteins exhibit distinct chromatin binding patterns, presumably due to their interactions with various specificity factors. Here, we elucidate the molecular basis of the protein-protein interaction between the HP1 protein Rhino, a critical factor of the Drosophila piRNA pathway, and Kipferl, a DNA sequence-specific C2H2 zinc finger protein and Rhino guidance factor. Through phylogenetic analyses, structure prediction, and in vivo genetics, we identify a single amino acid change within Rhino’s chromodomain, G31D, that does not affect H3K9me2/3 binding but abolishes the specific interaction between Rhino and Kipferl. Flies carrying the rhinoG31D mutation phenocopy kipferl mutant flies, with Rhino redistributing from piRNA clusters to satellite repeats, causing pronounced changes in the ovarian piRNA profile of rhinoG31D flies. Thus, Rhino’s chromodomain serves as a dual-specificity module, facilitating interactions with both a histone mark and a DNA-binding protein.

Introduction

Transposable elements (TEs) are ubiquitous in eukaryotic genomes but pose a significant threat to genome integrity (Bourque et al., 2018). When activated and mobile, these selfish genetic elements can lead to insertional mutagenesis and ectopic recombination events, imposing significant fitness costs on their hosts. To counteract the deleterious effects of TEs, eukaryotes package TE loci into repressive heterochromatin, effectively silencing these elements and preventing their uncontrolled movement within the genome (Levin and Moran, 2011, Fedoroff, 2012). Proteins of the heterochromatin protein 1 (HP1) family play a central role in the initiation and maintenance of heterochromatin from fungi to animals (Vermaak and Malik, 2009).

The founding member of the HP1 family, Drosophila Su(var)2-5, acts as a strong suppressor of position effect variegation (James and Elgin, 1986, Eissenberg et al., 1990, Eissenberg et al., 1992). It binds to heterochromatic histone marks and facilitates transcriptional silencing and the compaction of chromatin through the recruitment of histone methyltransferases, histone deacetylases, and other repressive activities (Vermaak and Malik, 2009, Allshire and Madhani, 2018). Most animal genomes encode multiple HP1 homologs that share a common domain architecture. They contain an N-terminal chromodomain with specific affinity for di- and tri-methylated histone H3 Lysine 9 (H3K9) peptides (Bannister et al., 2001, Lachner et al., 2001), an unstructured central hinge region of variable length involved in nonspecific nucleic acid interactions (Keller et al., 2012), and a C-terminal chromo shadow domain (Aasland and Stewart, 1995). While resembling the chromodomain fold, the chromo shadow domain does not bind histone tails. Instead, it forms a dimerization interface with the chromo shadow domain of another HP1 protein, creating a binding groove for proteins containing a PxV/LxL consensus motif (Smothers and Henikoff, 2000).

The number of HP1 family members varies between species. For instance, mice and humans encode three HP1 family proteins (HP1α, HP1β, HP1γ), whereas Drosophila melanogaster encodes five different members: the ubiquitously expressed HP1a/Su(var)2-5, HP1b, and HP1c proteins, and the germline-specific HP1d/Rhino (ovary and testis) and HP1e (testis) proteins (Vermaak and Malik, 2009, Levine et al., 2012). Despite having similar affinities for H3K9me2/3 reported from in vitro experiments, the Drosophila HP1 proteins have distinct biological functions and chromatin-binding patterns (Yu et al., 2015, Lee et al., 2019, Baumgartner et al., 2022). For example, while Su(var)2-5 binds all H3K9-methylated loci genome-wide, the germline-specific Rhino is enriched only at specific heterochromatic loci from where the non-coding precursors of PIWI-interacting RNAs (piRNAs) are transcribed (Vermaak et al., 2005, Klattenhoff et al., 2009, Mohn et al., 2014, Zhang et al., 2014). These so-called piRNA clusters are rich in repetitive sequences and serve as heritable sequence storage units that confer specificity to the piRNA pathway, a small RNA-based TE silencing system in animal gonads (Brennecke et al., 2007, Czech et al., 2018, Ozata et al., 2018). At the molecular level, Rhino facilitates the productive expression of heterochromatic piRNA clusters by recruiting specific effector proteins that stimulate transcription initiation, elongation, and nuclear export of the resulting non-coding piRNA precursors (Klattenhoff et al., 2009, Mohn et al., 2014, Zhang et al., 2014, Chen et al., 2016, Andersen et al., 2017, ElMaghraby et al., 2019, Kneuss et al., 2019). This makes Rhino a remarkably specialized HP1 protein that mediates activating rather than repressive chromatin identity. The tight regulation of Rhino’s chromatin-binding profile is therefore of great importance.

The zinc finger protein Kipferl, one of about ninety ZAD zinc finger proteins in Drosophila, acts as a critical guidance factor for Rhino in ovaries (Baumgartner et al., 2022). Kipferl binds to chromatin at genomic sites enriched in GRGGN motifs, presumably through a direct interaction between its C2H2 zinc finger arrays and DNA. When genomic Kipferl binding sites are located within an H3K9me2/3 domain, Kipferl recruits Rhino, and both proteins form extended binding domains around initial nucleation sites. The interaction between Kipferl and Rhino occurs between Kipferl’s fourth zinc finger and Rhino’s chromodomain. This interaction represents a highly unusual mode of binding because, unlike other interactions with HP1 proteins, it does not involve the dimeric HP1 chromo shadow domain.

Here, we reveal the molecular basis underlying the interaction between Kipferl and the Rhino chromodomain. We pinpointed a single amino acid adaptation within the chromodomain of Rhino that discriminates it from all other HP1 family members and that is critical for the specific Kipferl-Rhino interaction. Our findings provide important insights into how a direct protein-protein interaction governs the chromatin binding profile of an HP1 protein, illustrating how a single amino acid residue can contribute to the emergence of a novel protein function.

Results

Phylogenetic and structure prediction analyses of the Rhino-Kipferl interaction

Previous yeast two-hybrid (Y2H) experiments revealed a direct interaction between Kipferl and the chromodomain of Rhino, but not with that of the related Su(var)2-5 protein (Figure 1A) (Baumgartner et al., 2022). To investigate how the binding specificity of Kipferl for Rhino is ensured, we performed a comparative phylogenetic analysis of the chromodomains of Rhino, Su(var)2-5, HP1b, HP1c and HP1e homologs from different Drosophila species in which Kipferl orthologs could be unambiguously identified (Figure 1B; Figure 1 – figure supplement 1, Figure 1 – figure supplement 2). In this analysis, two residues stood out as Rhino-specific and conserved sequence alterations: the D31G mutation and the G62 insertion (Figure 1B).

Structure prediction and phylogenetic analyses point to a Rhino-specific residue involved in binding Kipferl.

(A) Domain organization of Kipferl and Rhino, with the AlphaFold pLDDT score plotted as a measure of order or disorder alongside. Red boxes indicate the smallest interacting fragments identified by yeast two-hybrid experiments by Baumgartner et al. (Baumgartner et al., 2022). ZAD, Zinc finger associated domain; ZnF, Zinc finger; CD, chromodomain; CSD, chromo shadow domain (B) Multiple sequence alignment of HP1 family proteins in five selected species harboring an unequivocally identified Kipferl homolog (see Figure 1 – figure supplement 3). Rhino-specific amino acid residues are indicated. Protein accessions and identifiers are documented in Supplementary File 1. Multi-Relief representation indicates residues that differ significantly in Rhino homologs versus other HP1 variant proteins. (C) PAE plot for the top ranked AlphaFold2 Multimer prediction of the Rhino chromodomain with the Kipferl ZnF cluster 1 (left) and structure of the complex in cartoon representation (Rhino, blue; Kipferl, green), together with the H3K9me3 peptide (orange) as observed in a Rhino – H3 crystal structure (PDB ID 4U68). Key residues of Rhino’s aromatic cage and H3K9me3, as well as of Kipferl’s C2H2 ZnF4 are shown in sticks representation. Only the interacting ZnF4 is shown. Depicted in the inset are Rhino G31 and HP1 D31, with HP1 (PDB ID 6MHA) superimposed on Rhino chromodomain residues 26-57 (RMSD = 0.55 Å), together with Kipferl V285 and F286, illustrating that D31 would lead to steric clashes with Kipferl.

To determine whether either of the two Rhino-specific residues could contribute to the interaction with Kipferl, we used AlphaFold2 Multimer (Jumper et al., 2021, Evans et al., 2022) to predict interactions between Rhino’s chromodomain and Kipferl’s first zinc finger array, which is composed of four C2H2 zinc fingers and was identified as interaction site with Rhino (Baumgartner et al., 2022). AlphaFold2 Multimer predicted a high confidence interaction with a single conformation in 5/5 models between both proteins, involving the fourth zinc finger of Kipferl, which was identified as necessary and sufficient for the Y2H interaction (Figure 1C, Figure 1 – figure supplement 3A, B, C) (Baumgartner et al., 2022). No interaction was predicted between Kipferl and the Su(var)2-5, HP1b, HP1c, or HP1e chromodomains. The predicted Kipferl-Rhino complex is compatible with binding to the H3K9me2/3 peptide through Rhino’s aromatic cage (Figure 1C) and would further allow for a potential dimerization of the Rhino chromodomain (Yu et al., 2015).

In the predicted complex, Kipferl’s fourth zinc finger interacts with Rhino’s chromodomain through an extended interface opposite of the aromatic cage, including β-sheets 2-4 and the C-terminal α-helix of the Rhino chromodomain. Centrally located in the predicted interface is the Rhino-specific glycine, G31, which occupies a position where all other HP1 proteins analyzed harbor a highly conserved aspartic acid (Figure 1B). Due to the nature of the predicted Kipferl-Rhino interaction, the exchange of glycine to aspartic acid in Rhino at position 31 would lead to steric clashes with the backbone and sidechains of Kipferl residues V285 and F286 and thus prevent the association of both proteins. We therefore hypothesized that mutating Rhino G31 to the HP1-typical aspartic acid residue (RhinoG31D) should specifically uncouple Rhino and Kipferl while leaving Rhino otherwise intact.

The RhinoG31D chromodomain retains H3K9me3 binding in vitro

Rhino’s in vivo function strictly depends on its ability to bind H3K9me2/3 via its chromodomain (Yu et al., 2015). In addition, it has been suggested that dimerization of the Rhino chromodomain is important for its function (Yu et al., 2015). To assess whether the Rhino G31D mutation impairs either of these functions, we analyzed a panel of recombinantly expressed Rhino chromodomains, including the wildtype construct, two putative Kipferl-binding mutants (G31A and G31D), and control mutants that impair H3K9me2/3 binding (mutations of the aromatic cage residues Y24A, W45A, or F48A) or putative dimerization (F34A/F76A double mutant) (Yu et al., 2015).

We used analytical size-exclusion chromatography with inline multi-angle light scattering (SEC-MALS) to assess the oligomeric state of the various Rhino chromodomain constructs. While our data replicated differences in elution volume between different mutant constructs (Yu et al., 2015), these did not correspond to substantial changes in their in-solution molecular weight, indicating that the oligomeric state remained the same for all constructs tested (Figure 2A; Figure 2 – figure supplement 1A). We conclude that the wildtype Rhino chromodomain, as well as the G31D or G31A variants, are monomeric in solution, as has been shown for other HP1 homologs (Jacobs et al., 2001, Brasher et al., 2000). To further test whether the G31D mutation leads to any unwanted structural changes in the Rhino chromodomain, we performed circular dichroism spectroscopy. All tested mutant constructs exhibited similar secondary structure composition compared to the wildtype construct (Figure 2 – figure supplement 1B).

G31 point mutations do not affect Rhino’s ability to bind H3K9me3.

(A) Line graph summarizing SEC-MALS results for the examined Rhino chromodomain constructs. The in solution molecular weight is indicated for each construct. (B) Isothermal titration calorimetry results showing the binding of indicated Rhino chromodomain constructs to the H3K9me3-modified histone tail peptide.

Having established that the two G31 mutant chromodomains do not exhibit altered protein folding or oligomeric state, we tested both constructs for their ability to bind H3K9me3 peptides alongside wildtype and aromatic cage mutant (F48A) controls. Consistent with previous observations (Yu et al., 2015, Le Thomas et al., 2014), isothermal titration calorimetry (ITC) experiments using synthetic H3K9me3 peptides revealed an affinity of 30.9 ± 3.0 μM for the wildtype domain and no measurable affinity for the F48A mutant (Figure 2B). Despite slight changes in the thermodynamic binding parameters, both the G31A and G31D mutants showed affinities comparable to the wildtype constructs with 43.5 ± 8.6 μM and 31.1 ± 3.2 μM, respectively. Thus, the RhinoG31D chromodomain behaves similarly to the wildtype construct in terms of oligomeric state, folding, and ability to bind H3K9me3 peptides in vitro.

The rhinoG31D mutant uncouples Rhino from Kipferl

To investigate the importance of G31 for Rhino function in vivo, we introduced a single point mutation into the endogenous rhino locus, reverting G31 to the aspartic acid found in all other HP1 proteins (rhinoG31D). In kipferl mutant females, Rhino fails to localize to most of its genomic binding sites, resulting in reduced piRNA levels and impaired fertility (Baumgartner et al., 2022). Females homozygous for the rhinoG31D allele were viable but exhibited severely reduced fertility: while the egg-laying rate of rhinoG31D females was comparable to that of control flies, the hatching rate of laid eggs dropped to 21 ± 9 % (Figure 3A). This decrease in fertility was less severe than the complete sterility seen for rhino null mutants, but similar to that observed in kipferl null mutants (Baumgartner et al., 2022), providing a first indication that the G31D mutation may specifically affect the Rhino–Kipferl interaction.

The rhinoG31Dpoint mutation recapitulates the mutant phenotypes observed for Rhino and Kipferl in each other’s null mutant background.

(A) Bar graph depicting female fertility as egg hatching rate in percent of laid eggs for indicated genotypes. (B) Confocal images showing immunofluorescence signal for Kipferl and Rhino in egg chambers of indicated genotypes. Zoomed images display one representative nurse cell nucleus (labeled by white asterisk in panel A) per genotype (scale bar: 20 µm).

To gain more direct insight into the Rhino–Kipferl interaction in rhinoG31D mutants, we examined changes in the typically pronounced colocalization of Kipferl and Rhino at discrete nuclear foci— corresponding to piRNA source loci—that is seen in wild-type nurse cells (Baumgartner et al., 2022). Using immunofluorescence imaging, we found a complete lack of colocalization of Rhino and Kipferl in rhinoG31D ovaries (Figure 3B). Kipferl localized diffusely in the nucleus with only a few foci, mirroring the localization of Kipferl in rhino null mutants. RhinoG31D was not enriched in these Kipferl foci, but instead accumulated in prominent structures at the nuclear envelope, reminiscent of Rhino accumulations found in kipferl null mutants (Baumgartner et al., 2022).

To determine the chromatin binding patterns of Rhino and Kipferl in ovaries of rhinoG31D mutant flies, we performed chromatin immunoprecipitation followed by sequencing (ChIP-seq). In wild-type ovaries, Rhino and Kipferl co-occupy hundreds of heterochromatic domains of different sizes in nearly identical enrichment patterns (Figure 4A) (Baumgartner et al., 2022). In addition, Kipferl, but not Rhino, binds to specific sites in euchromatin (Kipferl-only sites) that lack H3K9me2/3 marks but are enriched in GRGGN motifs, the presumed binding sequence of Kipferl. To account for the heterogeneous size of genomic Rhino/Kipferl domains, we analyzed their binding profiles by quantifying genome-unique ChIP-seq reads mapped to non-overlapping genomic 1 kilobase tiles (Mohn et al., 2014). In kipferl mutants, Rhino is lost from most of its genomic binding sites, with the tiles that retain Rhino binding primarily corresponding to piRNA clusters 38C and 42AB (Figure 4A, B) (Baumgartner et al., 2022). Conversely, in rhino mutants, Kipferl binding is still detected at euchromatic Kipferl-only sites, but it is strongly reduced at loci that are co-occupied by Kipferl and Rhino in wildtype: at sites where Rhino binding is Kipferl-dependent, Kipferl binding is reduced to more defined, narrow peaks. At Kipferl-independent loci on the other hand (e.g., piRNA clusters 38C and 42AB), Kipferl binding is almost completely lost in rhino mutants (Figure 4A). ChIP-seq experiments in rhinoG31D mutant ovaries revealed a chromatin occupancy for RhinoG31D that was almost indistinguishable from that of wild-type Rhino in kipferl mutants (Figure 4C, D). This extended to Kipferl-independent loci (e.g., piRNA clusters 38C and 42AB), where the slightly altered chromatin occupancy of Rhino in kipferl mutants was mirrored by RhinoG31D (Figure 4A). At the same time, the chromatin binding profile of Kipferl in rhinoG31D mutants strongly resembled that seen in rhino null-mutants genome-wide (Figure 4A, E). Taken together, mutation of a single Rhino-specific chromodomain residue to its ancestral state leads to the functional uncoupling of Rhino and Kipferl at the molecular level.

The RhinoG31D point mutation uncouples Rhino and Kipferl at chromatin.

(A) UCSC genome browser screenshots depicting the ChIP-seq signal for Rhino and Kipferl at diverse Rhino domains in ovaries of the indicated genotypes (signal shown as coverage per million sequenced reads for one representative replicate). (B-E) Scatter plot of genomic 1-kb tiles contrasting average log2-fold ChIP-seq enrichment for Rhino (B-D) or Kipferl (E) in ovaries of the indicated genotypes (values displayed represent the average of two to three replicate experiments)

RhinoG31D is functional at Kipferl-independent piRNA source loci

To determine whether the RhinoG31D point mutation affects the overall function of Rhino, we analyzed Kipferl-independent piRNA source loci. In kipferl mutant ovaries, Rhino is sequestered to large DNA satellite arrays, leading to greatly increased transcription and piRNA production at the Responder and 1.688 g/cm3 family satellites (Baumgartner et al., 2022). In rhinoG31D mutants, RNA fluorescent in situ hybridization (FISH) experiments showed that transcription of the Rsp and 1.688 g/cm3 satellites was also strongly elevated, resulting in elongated structures at the nuclear envelope, reminiscent of their phenotype observed in kipferl mutant nurse cell nuclei (Figure 5A). Consistent with their elevated transcription, RhinoG31D was enriched at satellite consensus sequences as determined by ChIP-seq, while it was reduced at most transposon sequences (Figure 5B, C). These observations extended to piRNA levels: piRNAs originating from Rsp and 1.688 g/cm3 satellites were strongly increased (Figure 5D), while piRNAs mapping to Rhino-dependent piRNA clusters were reduced specifically at Kipferl-dependent piRNA cluster 80F, but not at the Kipferl-independent piRNA clusters 38C and 42AB (Figure 5E). This further confirmed that the RhinoG31D mutation precisely phenocopies a kipferl null-mutant, indicating that RhinoG31D is fully functional at its remaining Kipferl-independent loci.

Kipferl-independent functions of Rhino are not affected by the G31D point mutation.

(A) Confocal images showing Rsp and 1.688 g/cm3 Satellite RNA FISH signal in nurse cells of indicated genotypes (scale bar: 5 µm). (B, C) Jitter plots depicting the log2-fold enrichments for Rhino ChIP-seq on consensus sequences of Satellites (B) or Rhino-dependent transposons (C) in ovaries with indicated genetic backgrounds. (D, G) Jitter plots depicting the length-normalized antisense piRNA counts on Satellite consensus sequences derived from ovaries (D) or testes (G) of indicated genetic backgrounds. (E, F) Box plots depicting the log2 fold change of piRNA counts (compared to w1118 control) per 1kb tile for major piRNA clusters in ovaries (E) or testes (F) of the indicated genotypes. The number of tiles per piRNA cluster is indicated as n.

Rhino also specifies dual-strand piRNA source loci in the male germline, where Kipferl is not expressed (Chen et al., 2021, Chen and Aravin, 2023, Baumgartner et al., 2022). piRNA source loci in testes only partially overlap with those of ovaries and are dynamically regulated during spermatogenesis. This implies Kipferl-independent mechanisms for Rhino recruitment to chromatin. To assess a potential impact of the G31D mutation on Rhino function in males, we sequenced small RNAs from testes issected from a panel of different genetic mutants. When rhino mutant testes were compared to wildtype controls, piRNA production from dual-strand piRNA source loci was impaired. In contrast, piRNA levels from the same loci remained essentially unchanged in kipferl or rhinoG31D mutants (Figure 5F). The same was observed for piRNAs derived from Rsp and 1.688 g/cm3 satellites, which are transcribed in a Rhino-dependent manner also in testes (Figure 5G). Taken together, the G31D mutation, while completely uncoupling Rhino from Kipferl, does not interfere with Rhino function at Kipferl-independent sites in ovaries and testes.

Conclusion

In this study we show how the DNA sequence-specific zinc finger protein, Kipferl, specifically binds the chromodomain of the HP1 variant Rhino, thereby dictating Rhino’s chromatin-binding profile and piRNA production in Drosophila. Our collective data show that reverting a single amino acid within Rhino’s chromodomain to its ancestral state, G31D, abolishes Kipferl’s ability to target Rhino to chromatin. Notably, the G31 residue in Rhino proteins is highly conserved among Drosophilids, even in species that lack a clearly identifiable Kipferl ortholog. This may indicate that other proteins use a mechanism similar to Kipferl to define Rhino’s chromatin occupancy in more distantly related Drosophila species. Our data also show that the RhinoG31D mutation does not affect the chromatin binding of Rhino at Kipferl-independent piRNA source loci in ovaries and testes, suggesting the existence of other, G31-independent recruitment mechanisms for Rhino to chromatin. Whether these alternative mechanisms act in a similar way to the one described here, utilizing zinc finger proteins and interactions with the Rhino chromodomain, remains to be determined. An important issue for future investigations, which is currently hampered by challenges in obtaining soluble recombinant Kipferl protein, will be to determine the precise three-dimensional arrangement of the Kipferl-Rhino complex together with Kipferl motif-containing DNA and H3K9-methylated nucleosomes, taking into account that Kipferl forms homodimers via its N-terminal ZAD domain and Rhino via its C-terminal chromo shadow domain.

Supporting information

eLife Assessment

The authors use a powerful combination of phylogenetics, structure prediction, biochemistry, and mutagenesis to provide an understanding of the mechanism that provides target specificity of Drosophila HP1 homolog Rhino vs. HP1, with Rhino specifically binding to piRNA loci. The authors show that a single amino acid substitution in the chromodomain of Rhino allows binding of the zinc finger protein Kipferl, which directs the complex to a subset of heterochromatic regions that other HP1 homologs do not. The evidence supporting the conclusions is compelling, providing an impressive level of mechanistic understanding of how the specificity of the piRNA genome defense system is defined. Also, the study highlights how a single amino acid change can change the functionality of a protein, providing fundamental insight into protein evolution.

Joint Public Review:

This article is a direct follow-up to the paper published last year in eLife by the same group. In the previous article, the authors discovered a zinc finger protein, Kipferl, capable of guiding the HP1 protein Rhino towards certain genomic regions enriched in GRGGN motifs and packaged in heterochromatin marked by H3K9me3. Unlike other HP1 proteins, Rhino recruitment activates the transcription of heterochromatic regions, which are then converted into piRNA source loci. The molecular mechanism by which Kipferl interacts specifically with Rhino (via its chromodomain) and not with other HP1 proteins remained enigmatic.

In this latest article, the authors go a step further by elucidating the molecular mechanisms important for the specific interaction of Rhino and not other HP1 proteins with Kipferl. A phylogenetic study carried out between the HP1 proteins of 5 Drosophila species led them to study the importance of an AA Glycine at position 31 located in the Rhino chromodomain, an AA different from the AA (aspartic acid) found at the same position in the other HP1 proteins. The authors then demonstrate, through a series of structure predictions, biochemical, and genetic experiments, that this specific AA in the Rhino-specific chromodomain explains the difference in the chromatin binding pattern between Rhino and the other Drosophila HP1 proteins. Importantly, the G31D conversion of the Rhino protein prevents interaction between Rhino and Kipferl, phenocopying a Kipfer mutant.

Strengths:

The authors' effective use of phylogenetic analyses and protein structure predictions to identify a substitution in the chromodomain that allows Rhino's specific interaction with Kipferl is very elegant. Both genetic and biochemical approaches are applied to rigorously probe the proposed explanation. They used a point mutation in the endogenous locus that replaces the Rhino-specific residue with the aspartic acid residue present in all other HP1 family members. This novel allele largely phenocopies the defects in hatch rate, chromatin organization, and piRNA production associated with kipferl mutants, and does not support Kipferl localization to clusters. The data are of high quality, the presentation is clear and concise, and the conclusions are generally well-supported.

Weaknesses:

The reviewers identified potential ways to further strengthen the manuscript.

1. The one significant omission is RNAseq on the rhino point mutant, which would allow direct comparison to cluster, transposon, and repeat expression in kipferl mutants.

2. The manuscript would benefit from adding more evolutionary comparisons. The following or similar analyses would help put the finding into a broader evolutionary perspective: i) Is Kipferl's surface interacting with Rhino also conserved in Kipferl orthologs? In other words, are the Rhino-interacting amino acids of Kipferl under any pressure to be conserved? ii) The remarkable conservation of Rhino's G31 is at odds with the arms race that is proposed to be happening between the fly's piRNA pathway proteins and transposons. Does this mean that Rhino's chromodomain is "untouchable" for such positive selection?

Author response:

The following is the authors’ response to the original reviews.

Public Review:

This article is a direct follow-up to the paper published last year in eLife by the same group. In the previous article, the authors discovered a zinc finger protein, Kipferl, capable of guiding the HP1 protein Rhino towards certain genomic regions enriched in GRGGN motifs and packaged in heterochromatin marked by H3K9me3. Unlike other HP1 proteins, Rhino recruitment activates the transcription of heterochromatic regions, which are then converted into piRNA source loci. The molecular mechanism by which Kipferl interacts specifically with Rhino (via its chromodomain) and not with other HP1 proteins remained enigmatic.

In this latest article, the authors go a step further by elucidating the molecular mechanisms important for the specific interaction of Rhino and not other HP1 proteins with Kipferl. A phylogenetic study carried out between the HP1 proteins of 5 Drosophila species led them to study the importance of an AA Glycine at position 31 located in the Rhino chromodomain, an AA different from the AA (aspartic acid) found at the same position in the other HP1 proteins. The authors then demonstrate, through a series of structure predictions, biochemical, and genetic experiments, that this specific AA in the Rhino-specific chromodomain explains the difference in the chromatin binding pattern between Rhino and the other Drosophila HP1 proteins. Importantly, the G31D conversion of the Rhino protein prevents interaction between Rhino and Kipferl, phenocopying a Kipferl mutant.

Strengths:

The authors' effective use of phylogenetic analyses and protein structure predictions to identify a substitution in the chromodomain that allows Rhino's specific interaction with Kipferl is very elegant. Both genetic and biochemical approaches are applied to rigorously probe the proposed explanation. They used a point mutation in the endogenous locus that replaces the Rhino-specific residue with the aspartic acid residue present in all other HP1 family members. This novel allele largely phenocopies the defects in hatch rate, chromatin organization, and piRNA production associated with kipferl mutants, and does not support Kipferl localization to clusters. The data are of high quality, the presentation is clear and concise, and the conclusions are generally well-supported.

Weaknesses:

The reviewers identified potential ways to further strengthen the manuscript.

(1) The one significant omission is RNAseq on the rhino point mutant, which would allow direct comparison to cluster, transposon, and repeat expression in kipferl mutants.

In this eLife Advances submission, we aim to elucidate the molecular interaction between Rhino and the zinc finger protein Kipferl and how it evolved. Using various assays, of which piRNA sequencing is the most relevant and comprehensive, we show that the rhino[G31D] mutation phenocopies a rhino loss-of-function situation for Kipferl and a kipferl loss-of-function situation for Rhino. Further confirmation of this statement by additional RNA-seq experiments to probe the extent of selective TE de-repression would indeed be a possibility. We decided to test for TE de-repression phenotypes using sensitive RNA-FISH experiments of a handful of TEs that are deregulated in kipferl loss of function flies (Baumgartner at al. 2022). This showed that the same TEs are also deregulated in rhino[G31D] flies, further confirming the similarity of the two genotypes. We have added these data to the text and to Figure 5-figure supplement 2, which shows representative RNA FISH images.

(2) The manuscript would benefit from adding more evolutionary comparisons. The following or similar analyses would help put the finding into a broader evolutionary perspective:

i) Is Kipferl's surface interacting with Rhino also conserved in Kipferl orthologs? In other words, are the Rhino-interacting amino acids of Kipferl under any pressure to be conserved?

We performed an analysis of the Kipferl interface that interacts with the Rhino chromodomain in those species where Kipferl could be unambiguously identified. This showed that the residues involved in the Rhino interaction are generally conserved. We have added this analysis to Figure 1-figure supplement 4.

ii) The remarkable conservation of Rhino's G31 is at odds with the arms race that is proposed to be happening between the fly's piRNA pathway proteins and transposons. Does this mean that Rhino's chromodomain is "untouchable" for such positive selection?

We agree that the conservation of the G31 residue argues against this binding interface being under positive selection in Rhino. Without understanding the pressures acting on Rhino that underlie the previously published positive selection, we find it difficult to draw firm conclusions. Mutating G31 in fly species that lack Kipferl would be an interesting experiment.

Recommendations for the authors:

(1) RNAseq is important to the full characterization of the phenotype and should be included. It's now clear that the major piRNA clusters are not required for fertility, so I would also include an analysis of piRNA production and Rhino binding to regions flanking isolated insertions.

See our response to raised weakness #1 above. Briefly, we have now added an analysis of TE de-repression based on RNA-FISH experiments (Figure 5-figure supplement 2). Regarding the proposed analysis of piRNA production and Rhino binding to regions flanking isolated TE insertions: this is an important issue that we carefully analysed in our previous work characterising the kipferl mutant (Baumgartner et al. 2022). In the present work, we focused on generating a rhino mutant that uncouples Rhino from Kipferl.

(2) The authors do not provide direct biochemical evidence that the chromodomain substitution blocks Rhino binding to Kipferl. However, Rhino protein is very low abundance, making analysis of the endogenous protein very difficult.

Based on our previous work (Baumgartner et al 2022), the Rhino chromodomain interacts directly with the fourth zinc finger of Kipferl. Mutation of a single residue in the predicted interface (Rhino[G31D]) phenocopies a kipferl mutant, strongly suggesting that this mutation disrupts the Rhino-Kipferl interaction. Definitive evidence will have to await the reconstitution of this interaction using recombinant proteins. Our attempts to purify recombinant Kipferl (expressed in bacteria or in insect cells) or the protein fragments relevant to the interaction were unsuccessful so far. While we obtained soluble fractions of the first ZnF array, there was always a high level of co-purifying nucleic acids that we were not able to remove.

(3) Even if the Rhino G31D mutant retains its ability to interact with H3K9me3 it does not localize correctly on the chromatin preventing certain regions such as locus 80F from being converted into piRNA source loci. However other regions such as satellite regions attract the Rhino mutant protein converting them into super piRNA source loci, phenocopying the effects observed in a Kipferl mutant. Why Rhino when not bound to Kipferl concentrates in satellite regions is a question that remains unanswered.

This is a very interesting question indeed. We have not been able to elucidate the molecular basis of how Rhino is recruited to satellite repeats in Kipferl mutants. For example, we performed a proximity biotinylation experiment with GFP-Rhino in Kipferl mutant ovaries, but this experiment did not reveal any protein that would explain the observed accumulation of Rhino at the complex satellite repeats.

(4) In the phylogenetic analysis the authors identified two residues as Rhino-specific and conserved sequence alterations, the D31G mutation and the G62 insertion. However, the authors limit their study to D31G mutation, and nothing is performed on the G32 insertion. It would have been interesting to know the impact of this insertion on Rhino's biology.

The role, if any, of the Rhino-specific G62 insertion and its effect on Rhino localisation or function is an interesting topic for further study. We have not investigated the G62 residue experimentally. In the current manuscript, we limited our efforts to the analysis of the G31D mutation, as the goal was to identify the mode of interaction with Kipferl, and the G62 residue is not predicted to contact Kipferl according to AlphaFold.

(5) The authors report that the G31D mutation of Rhino phenocopies the Kipferl mutant. Rhino is wrongly localized in the nucleus, and Rhino G31D recruitment in certain Kipferl-enriched regions is affected, as at the 80F locus, which correlates with a strong drop in piRNA production from this locus. To go a step further in demonstrating that G31D phenocopies the Kipferl mutant, it would have been informative to analyse how much TE piRNAs are affected and whether TEs are deregulated.

See our response to similar comments above. We have added RNA-FISH experiments to illustrate that the TE de-repression phenotypes are comparable between rhino[G31D] and kipferl loss of function ovaries (Figure 5-figure supplement 2). Analyses of TE-mapping piRNAs also show well correlated phenotypes (Figure 5-figure supplement 1).

(6) Figure 3A: To homogenize with the immunostaining presented in Figure 3B, can the authors add on the bar graph depicting female fertility the results obtained with kipferl-/- and rhino-/- genotype?

rhino mutants are completely (100%) sterile and the fertility of kipferl mutants was previously measured to range between 15% and 40% (Baumgartner et al. 2022).

(7) Figure 4A: It would have been interesting to show Venn diagrams showing the overlap of genomic regions enriched for Kipferl versus regions enriched for Rhi in a WT and in a Rhi G31D mutant.

We consider the analysis presented in Figure 4 to be more meaningful, as a Venn diagram would require binary cut-offs.

(8) Figure 1B: In the phylogenic analysis for Rhino/HP1d two D. simulans lines are presented. Can the authors clarify this point?

There are two Rhino paralogs in D. simulans: one paralog (NCBI: AAY34025.1) is more similar to D. melanogaster Rhino, contains one intron and is located at chromosome chr2R (assembly Apr. 2005, WUGSC mosaic 1.0/droSim1: 12256895-12258668). The second paralog (XP_002106478.1) is located on chromosome X (6734493-6735248) and does not contain an intron. We have added a clarifying statement to the corresponding figure legend.

(9) To determine whether Rhino G31D point mutation affects the overall function of Rhino, the authors analysed Kipferl-independent piRNA source loci by looking at Responder and 1,688 family satellites. I'm not sure that these loci can be classified as Kipferl-independent piRNA source loci since a strong increase of piRNA production from these loci in Kipferl mutant is observed. In my point of view, the 42AB and 38C are real Kipferl-independent piRNA source loci as piRNA production from these loci is not affected by Kipferl KD.

Indeed, the Rsp and 1,688 family satellites are not completely independent of Kipferl, as their expression and Rhino occupancy differ between wild-type and kipferl loss-of-function phenotypes (including rhino[G31D]). However, we believe that this increase is due to a strong dependence on different sequestration mechanisms and is not mediated by a direct function of Kipferl at these sites. Similarly, we observe slight differences in piRNA production for the peripheral parts of cluster 42AB, as well as differences in Rhino occupancy despite an unaltered piRNA profile at cluster 38C (Baumgartner et al. 2022). Thus, different flavours of Kipferl-independence exist, with the only truly Kipferl-independent piRNA sources likely to be the piRNA clusters in the testis. A clear classification is further complicated by previously observed compensatory effects in the piRNA pathway, leading us to adopt the current definition of "requiring Kipferl for Rhino recruitment" to distinguish Kipferl-dependent from Kipferl-independent sites.

(10) The authors report that the G31D mutation of Rhino phenocopies the Kipferl mutant. Rhino is wrongly localized in the nucleus, and Rhino G31D recruitment in certain Kipferl-enriched regions is affected, as at 80F locus, which correlates with a strong drop in piRNA production from this locus. To go a step further in demonstrating that G31D phenocopies the Kipferl mutant, it would have been interesting to look at how much TE piRNAs are affected and whether TEs (and which class of TE) are deregulated by RNAseq and/or in situ hybridization.

See our response to similar comments above. Our new RNA-FISH experiments and TE-mapping piRNA analysis extend the comparison of phenotypes between kipferl mutants and rhino[G31D] mutants and are consistent with our previous conclusions (Figure 5-figure supplements 1 and 2).

Version 2

Abstract

Abstract

Members of the diverse heterochromatin protein 1 (HP1) family play crucial roles in heterochromatin formation and maintenance. Despite the similar affinities of their chromodomains for di- and tri-methylated histone H3 lysine 9 (H3K9me2/3), different HP1 proteins exhibit distinct chromatin binding patterns, likely due to interactions with various specificity factors. Here, we elucidate the molecular basis of the interaction between the HP1 protein Rhino, a crucial factor of the Drosophila piRNA pathway, and Kipferl, a DNA sequence-specific C2H2 zinc finger protein and Rhino guidance factor. Through phylogenetic analyses, structure prediction, and in vivo genetics, we identify a single amino acid change within Rhino’s chromodomain, G31D, that does not affect H3K9me2/3 binding but disrupts the interaction between Rhino and Kipferl. Flies carrying the rhinoG31Dmutation phenocopy kipferl mutant flies, with Rhino redistributing from piRNA clusters to satellite repeats, causing pronounced changes in the ovarian piRNA profile of rhinoG31D flies. Thus, Rhino’s chromodomain functions as a dual-specificity module, facilitating interactions with both a histone mark and a DNA-binding protein.

Introduction

Transposable elements (TEs) are ubiquitous in eukaryotic genomes but pose a significant threat to genome integrity (Bourque et al., 2018). When activated and mobile, these selfish genetic elements can lead to insertional mutagenesis and ectopic recombination events, imposing significant fitness costs on their hosts. To counteract the deleterious effects of TEs, eukaryotes package TE loci into repressive heterochromatin, effectively silencing these elements and preventing their uncontrolled movement within the genome (Levin and Moran, 2011, Fedoroff, 2012). Proteins of the heterochromatin protein 1 (HP1) family play a central role in the initiation and maintenance of heterochromatin from fungi to animals (Vermaak and Malik, 2009).

The founding member of the HP1 family, Drosophila Su(var)2-5, acts as a strong suppressor of position effect variegation (James and Elgin, 1986, Eissenberg et al., 1990, Eissenberg et al., 1992). It binds to heterochromatic histone marks and facilitates transcriptional silencing and the compaction of chromatin through the recruitment of histone methyltransferases, histone deacetylases, and other repressive activities (Vermaak and Malik, 2009, Allshire and Madhani, 2018). Most animal genomes encode multiple HP1 homologs that share a common domain architecture. They contain an N-terminal chromodomain with specific affinity for di- and tri-methylated histone H3 Lysine 9 (H3K9) peptides (Bannister et al., 2001, Lachner et al., 2001), an unstructured central hinge region of variable length involved in nonspecific nucleic acid interactions (Keller et al., 2012), and a C-terminal chromo shadow domain (Aasland and Stewart, 1995). While resembling the chromodomain fold, the chromo shadow domain does not bind histone tails. Instead, it forms a dimerization interface with the chromo shadow domain of another HP1 protein, creating a binding groove for proteins containing a PxV/LxL consensus motif (Smothers and Henikoff, 2000).

The number of HP1 family members varies between species. For instance, mice and humans encode three HP1 family proteins (HP1α, HP1α, HP1γ), whereas Drosophila melanogaster encodes five different members: the ubiquitously expressed HP1a/Su(var)2-5, HP1b, and HP1c proteins, and the germline-specific HP1d/Rhino (ovary and testis) and HP1e (testis) proteins (Vermaak and Malik, 2009, Levine et al., 2012). Despite having similar affinities for H3K9me2/3 reported from in vitro experiments, the Drosophila HP1 proteins have distinct biological functions and chromatin-binding patterns (Yu et al., 2015, Lee et al., 2019, Baumgartner et al., 2022). For example, while Su(var)2-5 binds all H3K9-methylated loci genome-wide, the germline-specific Rhino is enriched only at specific heterochromatic loci from where the non-coding precursors of PIWI-interacting RNAs (piRNAs) are transcribed (Vermaak et al., 2005, Klattenhoff et al., 2009, Mohn et al., 2014, Zhang et al., 2014). These so-called piRNA clusters are rich in repetitive sequences and serve as heritable sequence storage units that confer specificity to the piRNA pathway, a small RNA-based TE silencing system in animal gonads (Brennecke et al., 2007, Czech et al., 2018, Ozata et al., 2018). At the molecular level, Rhino facilitates the productive expression of heterochromatic piRNA clusters by recruiting specific effector proteins that stimulate transcription initiation, elongation, and nuclear export of the resulting non-coding piRNA precursors (Klattenhoff et al., 2009, Mohn et al., 2014, Zhang et al., 2014, Chen et al., 2016, Andersen et al., 2017, ElMaghraby et al., 2019, Kneuss et al., 2019). This makes Rhino a remarkably specialized HP1 protein that mediates activating, rather than repressive, chromatin identity. The precise regulation of Rhino’s chromatin-binding profile is therefore of great importance.

The zinc finger protein Kipferl, one of about ninety ZAD zinc finger proteins in Drosophila, acts as a critical guidance factor for Rhino in ovaries (Baumgartner et al., 2022). Kipferl binds to chromatin at genomic sites enriched in GRGGN motifs, presumably through a direct interaction between its C2H2 zinc finger arrays and DNA. When genomic Kipferl binding sites are located within an H3K9me2/3 domain, Kipferl recruits Rhino, and both proteins form extended binding domains around initial nucleation sites. The interaction between Kipferl and Rhino occurs between Kipferl’s fourth zinc finger and Rhino’s chromodomain. This interaction represents a highly unusual mode of binding because, unlike other interactions with HP1 proteins, it does not involve the dimeric HP1 chromo shadow domain.

Here, we reveal the molecular basis underlying the interaction between Kipferl and the Rhino chromodomain. We identified a single amino acid adaptation within Rhino’s chromodomain that discriminates it from other HP1 family members and is critical for the specific Kipferl-Rhino interaction. Our findings provide important insights into how a direct protein-protein interaction dictates the chromatin binding profile of an HP1 protein, demonstrating how a single amino acid residue can contribute to the emergence of a novel protein function.

Results

Phylogenetic and structure prediction analyses of the Rhino-Kipferl interaction

Previous yeast two-hybrid (Y2H) experiments demonstrated a direct interaction between Kipferl and the chromodomain of Rhino, but not with that of the related Su(var)2-5 protein (Figure 1A) (Baumgartner et al., 2022). To explore the binding specificity of Kipferl for Rhino, we conducted a comparative phylogenetic analysis of the chromodomains of Rhino, Su(var)2-5, HP1b, HP1c and HP1e homologs from various Drosophila species with clearly identified Kipferl orthologs (Figure 1B; Figure 1 – figure supplement 1, Figure 1 – figure supplement 2). This analysis highlighted two Rhino-specific and conserved sequence alterations: the D31G change and the G62 insertion (Figure 1B).

Structure prediction and phylogenetic analyses point to a Rhino-specific residue involved in binding Kipferl.

(A) Domain organization of Kipferl and Rhino, with the AlphaFold pLDDT score plotted as a measure of order or disorder alongside. Red boxes indicate the smallest interacting fragments identified by yeast two-hybrid experiments by Baumgartner et al. (Baumgartner et al., 2022). ZAD, Zinc finger associated domain; ZnF, Zinc finger; CD, chromodomain; CSD, chromo shadow domain (B) Multiple sequence alignment of HP1 family proteins in five selected species harboring an unequivocally identified Kipferl homolog (see Figure 1 – figure supplement 2). Rhino-specific amino acid residues are indicated. Protein accessions and identifiers are documented in Supplementary File 1. Multi-Relief representation indicates residues that differ significantly in Rhino homologs versus other HP1 variant proteins. Note that two Rhino paralogs are identified in D. simulans (see Supplementary File 1 for accessions). (C) PAE plot for the top ranked AlphaFold2 Multimer prediction of the Rhino chromodomain with the Kipferl ZnF cluster 1 (left) and structure of the complex in cartoon representation (Rhino in blue; Kipferl in green), together with the H3K9me3 peptide (orange) as observed in a Rhino–H3K9me3 crystal structure (PDB ID 4U68). Key residues of Rhino’s aromatic cage and H3K9me3, as well as of Kipferl’s C2H2 ZnF4 are shown in sticks representation. Only the interacting ZnF4 is shown. Depicted in the inset are Rhino G31 and HP1 D31, with HP1 (PDB ID 6MHA) superimposed on Rhino chromodomain residues 26-57 (RMSD = 0.55 Å), together with Kipferl V285 and F286, illustrating that D31 would lead to steric clashes with Kipferl.

To explore whether either of the two Rhino-specific residues might contribute to the interaction with Kipferl, we used AlphaFold2 Multimer (Jumper et al., 2021, Evans et al., 2022) to predict interactions between Rhino’s chromodomain and Kipferl’s first zinc finger array, which comprises four C2H2 zinc fingers and was identified as the interaction site with Rhino (Baumgartner et al., 2022). AlphaFold2 Multimer predicted a high confidence interaction with a single conformation in 5/5 models, involving the fourth zinc finger of Kipferl, which is necessary and sufficient for the Y2H interaction with Rhino (Figure 1C, Figure 1 – figure supplement 3A, B, C) (Baumgartner et al., 2022). No interaction was predicted between Kipferl and the chromodomains of Su(var)2-5, HP1b, HP1c, or HP1e. The predicted Kipferl-Rhino complex is compatible with binding to the H3K9me2/3 peptide through Rhino’s aromatic cage (Figure 1C) and would allow for a potential dimerization of the Rhino chromodomain (Yu et al., 2015).

In the predicted complex, Kipferl’s fourth zinc finger interacts with Rhino’s chromodomain through an extended interface opposite the aromatic cage, including α-sheets 2-4 and the C-terminal α-helix of Rhino’s chromodomain (Figure 1C, Figure 1 – figure supplements 1 and 4). While the Rhino-specific G62 insertion does not participate in contacts with Kipferl, the Rhino-specific G31 residue, which in other HP1 proteins is a highly conserved aspartic acid, is centrally located in the predicted interaction interface (Figure 1C, Figure 1 – figure supplement 1). Due to the nature of the predicted Kipferl-Rhino interaction, substituting glycine with aspartic acid at position 31 in Rhino would cause steric clashes with Kipferl residues V285 and F286, preventing the association of both proteins. We therefore hypothesized that mutating Rhino G31 to the HP1-typical aspartic acid residue (RhinoG31D) should specifically uncouple Rhino and Kipferl while leaving Rhino otherwise functionally intact.

The RhinoG31D chromodomain retains H3K9me3 binding in vitro

Rhino’s in vivo function depends critically on its ability to bind H3K9me2/3 via its chromodomain (Yu et al., 2015). In addition, dimerization of the Rhino chromodomain has been suggested to be important for its function (Yu et al., 2015). To determine whether the Rhino G31D mutation affects either of these functions, we analyzed a panel of recombinantly expressed Rhino chromodomains. This panel included the wildtype construct, two putative Kipferl-binding mutants (G31A and G31D), and control mutants that impair H3K9me2/3 binding (mutations of the aromatic cage residues Y24A, W45A, or F48A) or putative dimerization (F34A/F76A double mutant) (Yu et al., 2015).

We used analytical size-exclusion chromatography with inline multi-angle light scattering (SEC-MALS) to assess the oligomeric state of the different Rhino chromodomain constructs. Our data confirmed differences in elution volume among the different mutant constructs (Yu et al., 2015), but these differences did not correspond to significant changes in their in-solution molecular weight, indicating that the oligomeric state remained consistent across all constructs tested (Figure 2A; Figure 2 – figure supplement 1). We conclude that the isolated wildtype Rhino chromodomain, along with the G31D or G31A variants, are monomeric in solution, as has been shown for other HP1 homologs (Jacobs et al., 2001, Brasher et al., 2000). To further investigate whether the G31D mutation causes any unwanted structural changes in the Rhino chromodomain, we performed circular dichroism spectroscopy. All tested mutant constructs exhibited similar secondary structure compositions compared to the wildtype construct (Figure 2 – figure supplement 2).

Rhino G31 point mutations do not affect Rhino’s ability to bind H3K9me3.

(A) Line graph summarizing SEC-MALS results for the examined Rhino chromodomain constructs. The in solution molecular weight is indicated for each construct. (B) Isothermal titration calorimetry results showing the binding of indicated Rhino chromodomain constructs to the H3K9me3-modified histone tail peptide.

Having established that the two G31 mutant chromodomains do not exhibit altered protein folding or oligomeric state, we tested both constructs for their ability to bind H3K9me3 peptides alongside wildtype and aromatic cage mutant (F48A) controls. Consistent with previous observations (Yu et al., 2015, Le Thomas et al., 2014), isothermal titration calorimetry (ITC) experiments using synthetic H3K9me3 peptides revealed an affinity of 30.9 ± 3.0 μM for the wildtype domain and no measurable affinity for the F48A mutant (Figure 2B). Despite slight changes in the thermodynamic binding parameters, both the G31A and G31D mutants showed affinities comparable to the wildtype constructs with 43.5 ± 8.6 μM and 31.1 ± 3.2 μM, respectively. Thus, the RhinoG31D chromodomain behaves similarly to the wildtype domain in terms of oligomeric state, folding, and ability to bind H3K9me3 peptides in vitro.

The rhinoG31D mutant uncouples Rhino from Kipferl

To explore the importance of G31 for Rhino function in vivo, we engineered a single point mutation within the endogenous rhino locus, converting G31 to the aspartic acid residue typically present in all other HP1 proteins (rhinoG31D). In kipferl mutant females, Rhino fails to localize to the majority of its genomic binding sites, resulting in diminished piRNA levels and compromised fertility (Baumgartner et al., 2022). Homozygous females carrying the rhinoG31D allele were viable but exhibited severely reduced fertility: Although the egg-laying rate of rhinoG31Dfemales was comparable to that of control flies, the hatching rate of laid eggs dropped to 21 ± 9 % (Figure 3A). While this decline in fertility was less severe compared to the complete sterility observed in rhino null mutants, it closely mirrored the impaired fertility of kipferl null mutants, which was in the range of 15 to 40% (Baumgartner et al., 2022), providing a first indication that the G31D mutation may specifically affect the Rhino–Kipferl interaction.

The rhinoG31D point mutation recapitulates the phenotypes for Rhino and Kipferl in each other’s null mutant background.

(A) Bar graph depicting female fertility as egg hatching rate in percent of laid eggs for indicated genotypes. (B) Confocal images showing immunofluorescence signal for Kipferl and Rhino in egg chambers of indicated genotypes. Zoomed images display one representative nurse cell nucleus (labeled by white asterisk in panel A) per genotype (scale bar: 20 µm).

To gain deeper insights into the Rhino–Kipferl interaction in rhinoG31Dmutants, we examined changes to the pronounced colocalization of Kipferl and Rhino at discrete nuclear foci – corresponding to piRNA source loci – observed in wild-type nurse cells (Baumgartner et al., 2022). Using immunofluorescence imaging, we observed a complete absence of colocalization between Rhino and Kipferl in rhinoG31Dmutants (Figure 3B). Kipferl localized diffusely in the nucleus with only a few foci, mirroring its distribution in rhino null mutants. RhinoG31D was not enriched within these Kipferl foci; instead, it accumulated in prominent structures near the nuclear envelope, resembling the Rhino accumulations found in kipferl null mutants (Baumgartner et al., 2022).

To determine the chromatin binding patterns of Rhino and Kipferl in ovaries of rhinoG31D mutant flies, we performed chromatin immunoprecipitation followed by sequencing (ChIP-seq). In wild-type ovaries, Rhino and Kipferl co-occupy hundreds of heterochromatic domains, displaying nearly identical enrichment patterns (Figure 4A) (Baumgartner et al., 2022). In addition, Kipferl, but not Rhino, binds to specific sites in euchromatin (Kipferl-only sites) that lack H3K9me2/3 marks but are enriched in GRGGN motifs, Kipferl’s presumed DNA binding motif. To account for the heterogeneous size of genomic Rhino/Kipferl domains, we analyzed their binding profiles by quantifying genome-unique ChIP-seq reads mapped to non-overlapping genomic 1-kilobase tiles (Mohn et al., 2014). In kipferl mutants, Rhino is lost from most of its genomic binding sites, with retained Rhino binding primarily corresponding to piRNA clusters 38C and 42AB (Figure 4A, B) (Baumgartner et al., 2022). Conversely, in rhino mutants, Kipferl binding persists at euchromatic Kipferl-only sites but is strongly reduced at loci that are co-occupied by Kipferl and Rhino in wildtype: at sites where Rhino binding is Kipferl-dependent, Kipferl binding is reduced to more defined, narrow peaks. At Kipferl-independent loci on the other hand (e.g., piRNA clusters 38C and 42AB), Kipferl binding is almost completely lost in rhino mutants (Figure 4A). ChIP-seq experiments in rhinoG31D mutant ovaries revealed a chromatin occupancy for RhinoG31D that was almost indistinguishable from that of wild-type Rhino in kipferl mutants (Figure 4C, D). This similarity extended to Kipferl-independent loci (e.g., piRNA clusters 38C and 42AB), where the altered chromatin occupancy of Rhino in kipferl mutants was mirrored by RhinoG31D (Figure 4A). At the same time, the chromatin binding profile of Kipferl in rhinoG31D mutants strongly resembled that observed in rhino null-mutants genome-wide (Figure 4A, E). Taken together, the mutation of a single Rhino-specific chromodomain residue to its ancestral state results in the functional uncoupling of Rhino and Kipferl at the molecular level.

The RhinoG31D point mutation uncouples Rhino and Kipferl on chromatin.

(A) UCSC genome browser screenshots depicting the ChIP-seq signal for Rhino and Kipferl at diverse Rhino domains in ovaries of the indicated genotypes (signal shown as coverage per million sequenced reads for one representative replicate). (B-E) Scatter plot of genomic 1-kb tiles contrasting average log2-fold ChIP-seq enrichment for Rhino (B-D) or Kipferl (E) in ovaries of the indicated genotypes (values displayed represent the average of two to three replicate experiments).

RhinoG31D is functional at Kipferl-independent piRNA source loci

To assess the impact of the RhinoG31D point mutation on Rhino’s overall functionality, we analyzed Kipferl-independent but Rhino-dependent piRNA source loci. In kipferl mutant ovaries, Rhino is sequestered to large DNA satellite arrays, resulting in greatly increased transcription and piRNA production at the Responder and 1.688 g/cm3 family satellites (Baumgartner et al., 2022). In rhinoG31D mutants, RNA fluorescent in situ hybridization (FISH) experiments showed that transcription of the Rsp and 1.688 g/cm3 satellites was also strongly elevated, leading to elongated structures at the nuclear envelope, reminiscent of the phenotype observed in kipferl mutant nurse cell nuclei (Figure 5A). Consistent with this elevated transcription, RhinoG31D was enriched at satellite consensus sequences as determined by ChIP-seq, while it was reduced at most transposon sequences (Figure 5B, C). These findings extended to piRNA levels: piRNAs originating from Rsp and 1.688 g/cm3 satellites were substantially increased (Figure 5D), while piRNAs were reduced at Kipferl-dependent piRNA clusters (e.g. cluster 80F), but not at Kipferl-independent piRNA clusters like 38C and 42AB (Figure 5E). Similarly, the levels of piRNAs mapping to transposon consensus sequences showed similar behaviors in rhinoG31D mutants as observed in kipferl mutants (Figure 5 – figure supplement 1A). This provides further confirmation that the RhinoG31D mutation faithfully phenocopies a kipferl null-mutant, indicating that RhinoG31D remains fully functional at Kipferl-independent loci. The altered piRNA levels observed in kipferl mutant ovaries result in the de-repression of a handful of transposable elements (Baumgartner et al., 2022). Based on RNA FISH experiments, the same transposons were also de-repressed in rhinoG31Dfemales, with the levels of upregulation resembling those in kipferl mutants rather than rhino mutants (Figure 5 – figure supplement 2), further suggesting a specific requirement of G31 for Kipferl-dependent functions of Rhino.

Kipferl-independent functions of Rhino are not affected by the G31D mutation.

(A) Confocal images showing Rsp and 1.688 g/cm3 Satellite RNA FISH signal in nurse cells of indicated genotypes (scale bar: 5 µm). (B, C) Jitter plots depicting the log2-fold enrichments for Rhino ChIP-seq on consensus sequences of Satellites (B) or Rhino-dependent transposons (C) in ovaries with indicated genetic backgrounds. (D, G) Jitter plots depicting the length-normalized antisense piRNA counts on Satellite consensus sequences derived from ovaries (D) or testes (G) of indicated genetic backgrounds. (E, F) Box plots depicting the log2 fold change of piRNA counts (compared to w1118 control) per 1kb tile for major piRNA clusters in ovaries (E) or testes (F) of the indicated genotypes. The number of tiles per piRNA cluster is indicated (n).

Rhino also plays a role in specifying piRNA source loci in the male germline, where Kipferl is not expressed (Chen et al., 2021, Chen and Aravin, 2023, Baumgartner et al., 2022). The piRNA source loci in testes only partially overlap with those of ovaries and are dynamically regulated during spermatogenesis, suggesting Kipferl-independent mechanisms for Rhino recruitment to chromatin. To assess a potential impact of the G31D mutation on Rhino function in males, we sequenced testes small RNAs from a panel of different genetic mutants. Comparing piRNAs from rhino mutant testes to wildtype controls confirmed the expected loss of piRNA production specifically from dual-strand piRNA source loci, whereas piRNA levels from the same loci remained unchanged in kipferl or rhinoG31Dmutants (Figure 5F). Consistent with this, the levels of transposon-mapping piRNAs also remained unaltered in kipferl or rhinoG31D mutants (Figure 5 – figure supplement 1B). The lack of a piRNA phenotype in testes further extended to the Rsp and 1.688 g/cm3 satellite loci, which produce Rhino-dependent piRNAs also in the male germline (Figure 5G). Taken together, the G31D mutation, while completely uncoupling Rhino from Kipferl, does not impede Rhino function at Kipferl-independent sites in either ovaries or testes.

Conclusion

In this study we elucidate the intricate interplay between the DNA sequence-specific zinc finger protein Kipferl, and the chromodomain of the HP1 variant Rhino. Our findings underscore the critical role of Kipferl in orchestrating Rhino’s chromatin-binding dynamics and subsequent piRNA production. Specifically, we show that a single amino acid alteration within Rhino’s chromodomain, reverting it to its ancestral state (G31D), disrupts Kipferl’s ability to target Rhino to chromatin. Notably, the G31 residue in Rhino is highly conserved among Drosophilids, even in species that lack a clearly identifiable Kipferl ortholog. This may indicate that other proteins use a mechanism similar to Kipferl to define Rhino’s chromatin occupancy in more distantly related Drosophila species. Our data also show that the RhinoG31D mutation does not affect the chromatin binding or the function of Rhino at Kipferl-independent piRNA source loci in ovaries and testes, suggesting the existence of other, G31-independent mechanisms for recruitment of Rhino to chromatin. Whether these alternative mechanisms act in a similar way to the one described here, utilizing zinc finger proteins and interactions with the Rhino chromodomain, remains an open question. An important issue for future investigation, currently hampered by the challenges of obtaining soluble recombinant Kipferl protein, will be to determine the precise three-dimensional arrangement of the Kipferl-Rhino complex together with Kipferl motif-containing DNA and H3K9-methylated nucleosomes, considering that Kipferl and Rhino are both likely to form homodimers via their N-terminal ZAD domain and C-terminal chromo shadow domain, respectively.

Supporting information